---
title: "Case study"
author: "James Zhan"
date: "2026-05-14"
output:
  html_document: default
  pdf_document: default
---

# Outline:
- Load packages and data
- Replace the missing intensity values with half minimum imputation
- Prepare the data for MSMICA targeted metabolomics analysis
- Prepare the data for MSMICA untargeted metabolomics analysis
- MSMICA targeted metabolomics analysis
    - Combine the MSMICA results from HILIC positive and C18 negative renal cancer study example data (7 paired renal cancer tissue samples and normal renal tissue samples taken from the same subjects) for a limma paired t-test comparison. Half minimum imputation is used for intensites with 0 values. 
        - Reference: https://bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf
- MSMICA untargeted metabolomics analysis
    - Using HILIC positive data and C18 negative data separately, perform a limma paired t-test comparison between renal cancer tissue and normal tissue samples within the same subject, select only features with FDR-adjusted p-value <= 0.20, and use MSMICA results to annotate the selected features. Half minimum imputation is used for intensites with 0 values.
- MSMICA targeted PLS Discriminant Analysis (PLS-DA)
    - Combine the HILIC positive and C18 negative MSMICA results for a PLS-DA analysis to compare renal cancer tissue and normal tissue samples within the same subject, followed by the sample grouping plot (shows the relationship and similaries between sample), hierarchical clustering heatmap, and relevance network analysis using mixOmics R package. Half minimum imputation is used for intensites with 0 values.
        - Reference: http://mixomics.org/case-studies/splsda-srbct-case-study/

---

## Load packages and data
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(limma)
library(splines)
library(MSMICA)

# this is the working directory for the case study
setwd('/Users/james/Desktop/Emory University - Ph.D./PhD dissertation/MSMICA/Publication/Abstract/MSMICA_package/MSMICA/vignettes')

# import feature table data
## hilic positive data
data(feature_table_exp_hilicpos)
print(feature_table_exp_hilicpos)
## c18 negative data
data(feature_table_exp_c18neg)
print(feature_table_exp_c18neg)

# replace all 0 values with NA after column 2
feature_table_exp_hilicpos[, -c(1:2)] = feature_table_exp_hilicpos[, -c(1:2)] %>%
    mutate_all(~na_if(., 0))

# replace all 0 values with NA after column 2
feature_table_exp_c18neg[, -c(1:2)] = feature_table_exp_c18neg[, -c(1:2)] %>%
    mutate_all(~na_if(., 0))

# median summarize the triplicates
df <- feature_table_exp_hilicpos

# Gather the triplicates into a long format
df_long <- df %>%
  pivot_longer(cols = -c(mz, time), names_to = "sample", values_to = "value")

# Extract the base sample name (without the replicate number)
df_long <- df_long %>%
  mutate(sample_base = sub("(_[0-9])$", "", sample))

# Compute the median for each set of triplicates
df_median <- df_long %>%
  group_by(mz, time, sample_base) %>%
  summarize(median_value = median(value), .groups = "drop")

# Pivot back to the wide format
feature_table_exp_hilicpos <- df_median %>%
  pivot_wider(names_from = sample_base, values_from = median_value)

# View the summarized data frame
print(feature_table_exp_hilicpos)

# median summarize the triplicates
df <- feature_table_exp_c18neg

# Gather the triplicates into a long format
df_long <- df %>%
  pivot_longer(cols = -c(mz, time), names_to = "sample", values_to = "value")

# Extract the base sample name (without the replicate number)
df_long <- df_long %>%
  mutate(sample_base = sub("(_[0-9])$", "", sample))

# Compute the median for each set of triplicates
df_median <- df_long %>%
  group_by(mz, time, sample_base) %>%
  summarize(median_value = median(value), .groups = "drop")

# Pivot back to the wide format
feature_table_exp_c18neg <- df_median %>%
  pivot_wider(names_from = sample_base, values_from = median_value)

# View the summarized data frame
print(feature_table_exp_c18neg)

# Filter out features appear less than 20% of all samples. The intensity column starts from the 3rd column.
feature_table_exp_hilicpos = QC_filter(x = feature_table_exp_hilicpos, metabolite_start_column = 3, minimum_sample_appear = 0.20)
feature_table_exp_c18neg = QC_filter(x = feature_table_exp_c18neg, metabolite_start_column = 3, minimum_sample_appear = 0.20)

# import MSMICA results
## hilic positive data
hilicpos_MSMICA_identified_metabolites = read_csv("MSMICA_test_hilicpos_MSMICA_identified_metabolites_filtered.csv")
print(hilicpos_MSMICA_identified_metabolites)
## c18 negative data
c18neg_MSMICA_identified_metabolites = read_csv("MSMICA_test_c18neg_MSMICA_identified_metabolites_filtered.csv")
print(c18neg_MSMICA_identified_metabolites)
```

## Replace the missing intensity values with half minimum imputation
```{r}

# Function to impute missing values with half of the minimum value in the row 
impute_missing_values <- function(row) {
  # Extract the values after column 2
  values <- row[-c(1, 2)]
  # Calculate half of the minimum value in the row (excluding NA)
  half_min_value <- min(values, na.rm = TRUE) / 2
  # Impute NA values with half of the minimum value
  values[is.na(values)] <- half_min_value
  # Combine the first two columns with the imputed values
  return(c(row[1:2], values))
}

# HILIC positive ############################################################################################################

# Apply the imputation function to each row
feature_table_exp_hilicpos_imputed <- t(apply(feature_table_exp_hilicpos, 1, impute_missing_values))

# Convert the result back to a data frame and ensure the column names are preserved
feature_table_exp_hilicpos_imputed <- as.data.frame(feature_table_exp_hilicpos_imputed, stringsAsFactors = FALSE)
colnames(feature_table_exp_hilicpos_imputed) <- colnames(feature_table_exp_hilicpos_imputed)
feature_table_exp_hilicpos_imputed = tibble(feature_table_exp_hilicpos_imputed)

# Round mz to 4 decimal places and time to 1 decimal place. This is because MSMICA already rounded mz to 4 decimal places and time to 0 decimal place. We need to match the feature table with MSMICA results later.
feature_table_exp_hilicpos_imputed$mz = round(feature_table_exp_hilicpos_imputed$mz, 4)
feature_table_exp_hilicpos_imputed$time = round(feature_table_exp_hilicpos_imputed$time, 0)

print(feature_table_exp_hilicpos_imputed)
# C18 negative ############################################################################################################

# Apply the imputation function to each row
feature_table_exp_c18neg_imputed <- t(apply(feature_table_exp_c18neg, 1, impute_missing_values))

# Convert the result back to a data frame and ensure the column names are preserved
feature_table_exp_c18neg_imputed <- as.data.frame(feature_table_exp_c18neg_imputed, stringsAsFactors = FALSE)
colnames(feature_table_exp_c18neg_imputed) <- colnames(feature_table_exp_c18neg_imputed)
feature_table_exp_c18neg_imputed = tibble(feature_table_exp_c18neg_imputed)

# Round mz to 4 decimal places and time to 1 decimal place. This is because MSMICA already rounded mz to 4 decimal places and time to 0 decimal place. We need to match the feature table with MSMICA results later.
feature_table_exp_c18neg_imputed$mz = round(feature_table_exp_c18neg_imputed$mz, 4)
feature_table_exp_c18neg_imputed$time = round(feature_table_exp_c18neg_imputed$time, 0)

print(feature_table_exp_c18neg_imputed)
```


## Prepare the data for MSMICA targeted metabolomics analysis
```{r}
# Prepare the data for MSMICA targeted metabolomics analysis ############################################################################################################

# to make sure there is no duplicate metabolites with the same name but multiple peaks, group by Name and select the metabolites with highest mean intensity
hilicpos_MSMICA_identified_metabolites = hilicpos_MSMICA_identified_metabolites %>%
    group_by(Name) %>%
    slice_max(mean_intensity) %>%
    ungroup()
c18neg_MSMICA_identified_metabolites = c18neg_MSMICA_identified_metabolites %>%
    group_by(Name) %>%
    slice_max(mean_intensity) %>%
    ungroup()

# transpose the table starting at column 32
hilicpos_MSMICA_featuretable_clean_t = t(hilicpos_MSMICA_identified_metabolites[, 32:ncol(hilicpos_MSMICA_identified_metabolites)])
c18neg_MSMICA_featuretable_clean_t = t(c18neg_MSMICA_identified_metabolites[, 32:ncol(c18neg_MSMICA_identified_metabolites)])

# convert the table to a data frame
hilicpos_MSMICA_featuretable_clean_t = as.data.frame(hilicpos_MSMICA_featuretable_clean_t)
c18neg_MSMICA_featuretable_clean_t = as.data.frame(c18neg_MSMICA_featuretable_clean_t)

# set the column names as the Identified_Name
colnames(hilicpos_MSMICA_featuretable_clean_t) = hilicpos_MSMICA_identified_metabolites$Name
colnames(c18neg_MSMICA_featuretable_clean_t) = c18neg_MSMICA_identified_metabolites$Name

# add Group_ID as a column
hilicpos_MSMICA_featuretable_clean_t$Group_ID = rownames(hilicpos_MSMICA_featuretable_clean_t)
c18neg_MSMICA_featuretable_clean_t$Group_ID = rownames(c18neg_MSMICA_featuretable_clean_t)

# move Group_ID as the first column
hilicpos_MSMICA_featuretable_clean_t = tibble(hilicpos_MSMICA_featuretable_clean_t) %>%
    select(Group_ID, everything())
c18neg_MSMICA_featuretable_clean_t = tibble(c18neg_MSMICA_featuretable_clean_t) %>%
    select(Group_ID, everything())

# Seperate the Group_ID into columns: Group, Temperature, Time, SampleID
hilicpos_MSMICA_featuretable_clean_t = hilicpos_MSMICA_featuretable_clean_t %>%
    separate(Group_ID, c("Group", "Temperature", "Time", "Triplicate"), sep = "_") %>%
    # create a new column, sample_tyoe, to indicate the sample type: if Group starts with NM, then it is normal, otherwise it is cancer
    mutate(sample_type = ifelse(stringr::str_detect(Group, "^NM"), "normal", "cancer")) %>%
    # move sample type to the first column
    select(sample_type, everything())
c18neg_MSMICA_featuretable_clean_t = c18neg_MSMICA_featuretable_clean_t %>%
    separate(Group_ID, c("Group", "Temperature", "Time", "Triplicate"), sep = "_") %>%
    # create a new column, sample_tyoe, to indicate the sample type: if Group starts with NM, then it is normal, otherwise it is cancer
    mutate(sample_type = ifelse(stringr::str_detect(Group, "^NM"), "normal", "cancer")) %>%
    # move sample type to the first column
    select(sample_type, everything())

# separate the Group column into Group and ParticipantID: the first two characters are Group, the rest are ParticipantID
hilicpos_MSMICA_featuretable_clean_t = hilicpos_MSMICA_featuretable_clean_t %>%
    separate(Group, c("Group", "ParticipantID"), sep = 2)
c18neg_MSMICA_featuretable_clean_t = c18neg_MSMICA_featuretable_clean_t %>%
    separate(Group, c("Group", "ParticipantID"), sep = 2)

# Convert all intensity values to log2
hilicpos_MSMICA_featuretable_clean_t[, -c(1:6)] = hilicpos_MSMICA_featuretable_clean_t[, -c(1:6)] %>%
    mutate_all(~log2(.))
c18neg_MSMICA_featuretable_clean_t[, -c(1:6)] = c18neg_MSMICA_featuretable_clean_t[, -c(1:6)] %>%
    mutate_all(~log2(.))

# check the data ready for MSMICA targeted metabolomics analysis
print(hilicpos_MSMICA_featuretable_clean_t)
print(c18neg_MSMICA_featuretable_clean_t)

# save the results as a csv file
write.csv(hilicpos_MSMICA_featuretable_clean_t, "MSMICA targeted metabolomics analysis/hilicpos_MSMICA_featuretable_clean_transposed.csv", row.names = FALSE)
write.csv(c18neg_MSMICA_featuretable_clean_t, "MSMICA targeted metabolomics analysis/c18neg_MSMICA_featuretable_clean_transposed.csv", row.names = FALSE)
```


## MSMICA targeted metabolomics analysis
```{r}
############################# limma paired t-test #############################
# Factorize the samples into normal and cancer
Group_hilicpos = factor(hilicpos_MSMICA_featuretable_clean_t$sample_type, levels = c("normal", "cancer"))
Group_c18neg = factor(c18neg_MSMICA_featuretable_clean_t$sample_type, levels = c("normal", "cancer"))

# Factorize the ParticipantID
ParticipantID_hilicpos = factor(hilicpos_MSMICA_featuretable_clean_t$ParticipantID)
ParticipantID_c18neg = factor(c18neg_MSMICA_featuretable_clean_t$ParticipantID)

# Create the design matrix
design_T_hilicpos <- model.matrix(~ ParticipantID_hilicpos+Group_hilicpos)
design_T_c18neg <- model.matrix(~ ParticipantID_c18neg+Group_c18neg)

# Extract the intensity values for the limma analysis
hilicpos_intensity = hilicpos_MSMICA_featuretable_clean_t[, 7:ncol(hilicpos_MSMICA_featuretable_clean_t)]
c18neg_intensity = c18neg_MSMICA_featuretable_clean_t[, 7:ncol(c18neg_MSMICA_featuretable_clean_t)]
hilicpos_intensity = as.data.frame(hilicpos_intensity)
c18neg_intensity = as.data.frame(c18neg_intensity)

# transpose the metabolite intensity data
y_T_hilicpos = t(hilicpos_intensity)
y_T_c18neg = t(c18neg_intensity)

# Fit the linear model
fit_hilicpos = lmFit(y_T_hilicpos, design_T_hilicpos)
fit_c18neg = lmFit(y_T_c18neg, design_T_c18neg)
fit2_hilicpos <- eBayes(fit_hilicpos)
fit2_c18neg <- eBayes(fit_c18neg)

# Extract all relevant values from the limma output
limma_outcome_hilicpos = tibble(
    Cancer_log2FC = c(fit2_hilicpos$coefficients[, "Group_hilicposcancer"]),
    t_statistic = c(fit2_hilicpos$t[, "Group_hilicposcancer"]), 
    p = c(fit2_hilicpos$p.value[, "Group_hilicposcancer"]))

limma_outcome_c18neg = tibble(
    Cancer_log2FC = c(fit2_c18neg$coefficients[, "Group_c18negcancer"]),
    t_statistic = c(fit2_c18neg$t[, "Group_c18negcancer"]), 
    p = c(fit2_c18neg$p.value[, "Group_c18negcancer"]))

limma_outcome_hilicpos$p.fdr = p.adjust(limma_outcome_hilicpos$p, method="BH")
limma_outcome_c18neg$p.fdr = p.adjust(limma_outcome_c18neg$p, method="BH")
limma_outcome_hilicpos$metabolite = colnames(hilicpos_MSMICA_featuretable_clean_t[, -c(1:6)])
limma_outcome_c18neg$metabolite = colnames(c18neg_MSMICA_featuretable_clean_t[, -c(1:6)])

# add Name, HMDB_ID, KEGG_ID, Mono_mass, Adduct, mz, time from hilicpos_MSMICA_identified_metabolites to limma_outcome_hilicpos
limma_outcome_hilicpos$Name = hilicpos_MSMICA_identified_metabolites$Name
limma_outcome_hilicpos$HMDB_ID = hilicpos_MSMICA_identified_metabolites$HMDB_ID
limma_outcome_hilicpos$KEGG_ID = hilicpos_MSMICA_identified_metabolites$KEGG_ID
limma_outcome_hilicpos$Mono_mass = hilicpos_MSMICA_identified_metabolites$Mono_mass
limma_outcome_hilicpos$Adduct = hilicpos_MSMICA_identified_metabolites$Adduct
limma_outcome_hilicpos$mz = hilicpos_MSMICA_identified_metabolites$mz
limma_outcome_hilicpos$time = hilicpos_MSMICA_identified_metabolites$time

limma_outcome_c18neg$Name = c18neg_MSMICA_identified_metabolites$Name
limma_outcome_c18neg$HMDB_ID = c18neg_MSMICA_identified_metabolites$HMDB_ID
limma_outcome_c18neg$KEGG_ID = c18neg_MSMICA_identified_metabolites$KEGG_ID
limma_outcome_c18neg$Mono_mass = c18neg_MSMICA_identified_metabolites$Mono_mass
limma_outcome_c18neg$Adduct = c18neg_MSMICA_identified_metabolites$Adduct
limma_outcome_c18neg$mz = c18neg_MSMICA_identified_metabolites$mz
limma_outcome_c18neg$time = c18neg_MSMICA_identified_metabolites$time

# Organize the column names: Name, HMDB_ID, KEGG_ID, Mono_mass, Adduct, mz, time, t_statistics, p, p.fdr, Cancer_log2FC
limma_outcome_hilicpos = limma_outcome_hilicpos %>%
    select(Name, HMDB_ID, KEGG_ID, Mono_mass, Adduct, mz, time, t_statistic, p, p.fdr, Cancer_log2FC) %>%
    arrange(p)

limma_outcome_c18neg = limma_outcome_c18neg %>%
    select(Name, HMDB_ID, KEGG_ID, Mono_mass, Adduct, mz, time, t_statistic, p, p.fdr, Cancer_log2FC) %>%
    arrange(p)

# select only the significant metabolites
limma_outcome_hilicpos_sig = limma_outcome_hilicpos %>%
    filter(p < 0.05)
limma_outcome_c18neg_sig = limma_outcome_c18neg %>%
    filter(p < 0.05)

# print the results
print(limma_outcome_hilicpos_sig, width=Inf)
print(limma_outcome_c18neg_sig, width=Inf)

# save the result as a csv file
write.csv(limma_outcome_hilicpos_sig, "MSMICA targeted metabolomics analysis/rcc_targeted_MSMICA_HILICpos_metabolites_limma_paired_t_test.csv", row.names = FALSE)
write.csv(limma_outcome_c18neg_sig, "MSMICA targeted metabolomics analysis/rcc_targeted_MSMICA_C18neg_metabolites_limma_paired_t_test.csv", row.names = FALSE)
```


## MSMICA targeted PLS Discriminant Analysis (PLS-DA)
```{r, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
library(readr)
library(dplyr)
library(tidyr)
library(mixOmics)

set.seed(2024) # for reproducibility, remove for normal use

# This is the intensity data for the MSMICA identified metabolites in HILIC positive only
X = as.matrix(hilicpos_intensity)

# This is the sample type for each sample: normal or cancer
Y = factor(hilicpos_MSMICA_featuretable_clean_t$sample_type, levels = c("normal", "cancer"))

dim(X) # check the dimensions of the X dataframe
length(Y) # check the length of the Y dataframe

# Preliminary Analysis with PCA
# run pca method on data (our intensity data was already log2 transformed, so we don't need to scale it here)
pca.RCC = pca(X, ncomp = 2, center = TRUE, scale = FALSE) 
plot(pca.RCC)  # barplot of the eigenvalues (explained variance per component)

# Two components would be sufficient to explain a moderate proportion of the data's variance
plotIndiv(pca.RCC, group = Y, ind.names = TRUE, # plot the samples projected
          legend = TRUE, 
          col.per.group = c("#F68B33", "#388ECC"), # set of colours
          title = 'PCA on the RCC study, comp 1 & 2',
          ellipse = TRUE) # onto the PCA

# It seems like only the normal tissue samples are separated from the cancer tissue samples. Cancer tissue samples are heterogenous. This is expected as the normal tissue samples are expected to have different metabolite profiles compared to the cancer tissue samples.

# PLS-DA - Note, PLS-DA is a supervised method, so we need to provide the class labels (Y) to the function. This is designed to find the metabolites that best discriminate between the two classes. Thus, it is not surprising that the PLS-DA model will perform better than the PCA model for class discrimination.
RCC_plsda <- plsda(X, Y, ncomp = 2)  # set ncomp to 2 for performance assessment later

# plot the samples projected onto the first two components of the PLS-DA subspace
plotIndiv(RCC_plsda, comp = c(1,2), # plot samples from final model
          group = Y, ind.names = TRUE, # colour by class label
          ellipse = TRUE, legend = TRUE, # include 95% confidence ellipse,
          col.per.group = c("#F68B33", "#388ECC"), # set of colours
          title = 'Final PLS-DA on the RCC study, comp 1 & 2')

# list their VIP scores
# Obtain the VIP (Variable Importance in Projection) scores for the fitted PLS-DA model
VIP_scores <- mixOmics::vip(RCC_plsda)

# extract the  VIP scores
VIP_scores = VIP_scores[, 1]
VIP_scores_df = tibble(
    Name = names(VIP_scores),
    VIP_scores = as.numeric(VIP_scores))
VIP_scores_df = VIP_scores_df %>%
    arrange(desc(VIP_scores))
print(VIP_scores_df)

# save the VIP scores as a csv file
write.csv(VIP_scores_df, "MSMICA targeted metabolomics analysis/rcc_targeted_MSMICA_HILICpos_metabolites_PLS-DA_VIP_scores.csv", row.names = FALSE)
```


## Prepare the data for MSMICA untargeted metabolomics analysis
```{r}
# Prepare the data for MSMICA untargeted metabolomics analysis ############################################################################################################
# HILIC positive ############################################################################################################
# all MSMICA identified metabolites in HILIC positive and C18 negative #############################
# transpose the table starting at column 13
feature_table_exp_hilicpos_imputed_t = t(feature_table_exp_hilicpos_imputed[,3:ncol(feature_table_exp_hilicpos_imputed)])
feature_table_exp_hilicpos_imputed_t = as.data.frame(feature_table_exp_hilicpos_imputed_t)

# set the column names as the combination of mz and time
colnames(feature_table_exp_hilicpos_imputed_t) = paste0(feature_table_exp_hilicpos_imputed$mz, "_", feature_table_exp_hilicpos_imputed$time)

# add Group_ID as a column
feature_table_exp_hilicpos_imputed_t$Group_ID = rownames(feature_table_exp_hilicpos_imputed_t)

# move Group_ID as the first column
feature_table_exp_hilicpos_imputed_t = tibble(feature_table_exp_hilicpos_imputed_t) %>%
    dplyr::select(Group_ID, everything())

# Seperate the Group_ID into columns: Group, Temperature, Time, SampleID
feature_table_exp_hilicpos_imputed_t = feature_table_exp_hilicpos_imputed_t %>%
    tidyr::separate(Group_ID, c("Group", "Temperature", "Time"), sep = "_") %>%
    # create a new column, sample_tyoe, to indicate the sample type: if Group starts with NM, then it is normal, otherwise it is cancer
    mutate(sample_type = ifelse(stringr::str_detect(Group, "^NM"), "normal", "cancer")) %>%
    # move sample type to the first column
    dplyr::select(sample_type, everything())
    
# separate the Group column into Group and ParticipantID: the first two characters are Group, the rest are ParticipantID
feature_table_exp_hilicpos_imputed_t = feature_table_exp_hilicpos_imputed_t %>%
    tidyr::separate(Group, c("Group", "ParticipantID"), sep = 2)

# Convert all intensity values to log2
feature_table_exp_hilicpos_imputed_t[, -c(1:5)] = feature_table_exp_hilicpos_imputed_t[, -c(1:5)] %>%
    mutate_all(~log2(.))

# check the data ready for MSMICA untargeted metabolomics analysis
print(feature_table_exp_hilicpos_imputed_t)

# C18 negative ############################################################################################################
# all MSMICA identified metabolites #############################
feature_table_exp_c18neg_imputed_t = t(feature_table_exp_c18neg_imputed[,3:ncol(feature_table_exp_c18neg_imputed)])
feature_table_exp_c18neg_imputed_t = as.data.frame(feature_table_exp_c18neg_imputed_t)

# set the column names as the combination of mz and time
colnames(feature_table_exp_c18neg_imputed_t) = paste0(feature_table_exp_c18neg_imputed$mz, "_", feature_table_exp_c18neg_imputed$time)

# add Group_ID as a column
feature_table_exp_c18neg_imputed_t$Group_ID = rownames(feature_table_exp_c18neg_imputed_t)

# move Group_ID as the first column
feature_table_exp_c18neg_imputed_t = tibble(feature_table_exp_c18neg_imputed_t) %>%
    dplyr::select(Group_ID, everything())

# Seperate the Group_ID into columns: Group, Temperature, Time, SampleID
feature_table_exp_c18neg_imputed_t = feature_table_exp_c18neg_imputed_t %>%
    separate(Group_ID, c("Group", "Temperature", "Time"), sep = "_") %>%
    # create a new column, sample_tyoe, to indicate the sample type: if Group starts with NM, then it is normal, otherwise it is cancer
    mutate(sample_type = ifelse(stringr::str_detect(Group, "^NM"), "normal", "cancer")) %>%
    # move sample type to the first column
    dplyr::select(sample_type, everything())

# separate the Group column into Group and ParticipantID: the first two characters are Group, the rest are ParticipantID
feature_table_exp_c18neg_imputed_t = feature_table_exp_c18neg_imputed_t %>%
    separate(Group, c("Group", "ParticipantID"), sep = 2)

# Convert all intensity values to log2
feature_table_exp_c18neg_imputed_t[, -c(1:5)] = feature_table_exp_c18neg_imputed_t[, -c(1:5)] %>%
    mutate_all(~log2(.))

# check the data ready for MSMICA untargeted metabolomics analysis
print(feature_table_exp_c18neg_imputed_t)
```


## MSMICA untargeted metabolomics analysis
```{r}
############################# limma paired t-test #############################
# HILIC positive ############################################################################################################
# Factorize the samples into normal and cancer
Group_hilicpos = factor(feature_table_exp_hilicpos_imputed_t$sample_type, levels = c("normal", "cancer"))

# Factorize the ParticipantID
ParticipantID_hilicpos = factor(feature_table_exp_hilicpos_imputed_t$ParticipantID)

# Create the design matrix
design_T_hilicpos <- model.matrix(~ ParticipantID_hilicpos+Group_hilicpos)

# Extract the intensity values for the limma analysis
hilicpos_intensity = feature_table_exp_hilicpos_imputed_t[, 6:ncol(feature_table_exp_hilicpos_imputed_t)]
hilicpos_intensity = as.data.frame(hilicpos_intensity)

# transpose the metabolite intensity data
y_T_hilicpos = t(hilicpos_intensity)

# Fit the linear model
fit_hilicpos = lmFit(y_T_hilicpos, design_T_hilicpos)
fit2_hilicpos <- eBayes(fit_hilicpos)

# chcek the top table in the limma paited t-test output
topTable(fit2_hilicpos, coef="Group_hilicposcancer", adjust="BH")

# Extract all relevant values from the limma output
limma_outcome_hilicpos = data.frame(
    Cancer_log2FC = c(fit2_hilicpos$coefficients[, "Group_hilicposcancer"]),
    t_statistic = c(fit2_hilicpos$t[, "Group_hilicposcancer"]), 
    p = c(fit2_hilicpos$p.value[, "Group_hilicposcancer"]))

limma_outcome_hilicpos$p.fdr = p.adjust(limma_outcome_hilicpos$p, method="BH")
limma_outcome_hilicpos$mz_time = colnames(feature_table_exp_hilicpos_imputed_t[, -c(1:5)])
limma_outcome_hilicpos = tibble(limma_outcome_hilicpos)

# mz, time from feature_table_exp_hilicpos_imputed to limma_outcome_hilicpos
limma_outcome_hilicpos$mz = feature_table_exp_hilicpos_imputed$mz
limma_outcome_hilicpos$time = feature_table_exp_hilicpos_imputed$time

# Organize the column names: mz, time, t_statistics, p, p.fdr, Cancer_log2FC
limma_outcome_hilicpos = limma_outcome_hilicpos %>%
    dplyr::select(mz, time, t_statistic, p, p.fdr, Cancer_log2FC) %>%
    arrange(p) %>%
    tibble()

# select only the significant metabolites
limma_outcome_hilicpos_sig_untargeted = limma_outcome_hilicpos %>%
    filter(p < 0.05)

# left join the MSMICA results to the limma_outcome_hilicpos_sig based on mz and time
limma_outcome_hilicpos_sig_untargeted_MSMICA = left_join(limma_outcome_hilicpos_sig_untargeted, hilicpos_MSMICA_identified_metabolites, by=c("mz"="mz", "time"="time"))

# print the results
print(limma_outcome_hilicpos_sig_untargeted_MSMICA)

# save the result as a csv file
write.csv(limma_outcome_hilicpos_sig_untargeted_MSMICA, "MSMICA untargeted metabolomics analysis/rcc_MSMICA_untargeted_hilicpos_metabolites_limma_paired_t_test.csv", row.names = FALSE)

# C18 negative ############################################################################################################
# Factorize the samples into normal and cancer
Group_c18neg = factor(feature_table_exp_c18neg_imputed_t$sample_type, levels = c("normal", "cancer"))

# Factorize the ParticipantID
ParticipantID_c18neg = factor(feature_table_exp_c18neg_imputed_t$ParticipantID)

# Create the design matrix
design_T_c18neg <- model.matrix(~ ParticipantID_c18neg+Group_c18neg)

# Extract the intensity values for the limma analysis
c18neg_intensity = feature_table_exp_c18neg_imputed_t[, 6:ncol(feature_table_exp_c18neg_imputed_t)]

# transpose the metabolite intensity data
y_T_c18neg = t(c18neg_intensity)

# Fit the linear model
fit_c18neg = lmFit(y_T_c18neg, design_T_c18neg)
fit2_c18neg <- eBayes(fit_c18neg)

# chcek the top table in the limma paited t-test output
topTable(fit2_c18neg, coef="Group_c18negcancer", adjust="BH")

# Extract all relevant values from the limma output
limma_outcome_c18neg = data.frame(
    Cancer_log2FC = c(fit2_c18neg$coefficients[, "Group_c18negcancer"]),
    t_statistic = c(fit2_c18neg$t[, "Group_c18negcancer"]), 
    p = c(fit2_c18neg$p.value[, "Group_c18negcancer"]))

limma_outcome_c18neg$p.fdr = p.adjust(limma_outcome_c18neg$p, method="BH")
limma_outcome_c18neg$mz_time = colnames(feature_table_exp_c18neg_imputed_t[, -c(1:5)])
limma_outcome_c18neg = tibble(limma_outcome_c18neg)

# mz, time from feature_table_exp_c18neg_imputed to limma_outcome_c18neg
limma_outcome_c18neg$mz = feature_table_exp_c18neg_imputed$mz
limma_outcome_c18neg$time = feature_table_exp_c18neg_imputed$time

# Organize the column names: mz, time, t_statistics, p, p.fdr, Cancer_log2FC
limma_outcome_c18neg = limma_outcome_c18neg %>%
    dplyr::select(mz, time, t_statistic, p, p.fdr, Cancer_log2FC) %>%
    arrange(p) %>%
    tibble()

# select only the significant metabolites
limma_outcome_c18neg_sig_untargeted = limma_outcome_c18neg %>%
    filter(p < 0.05)

# left join the MSMICA results to the limma_outcome_c18neg_sig based on mz and time
limma_outcome_c18neg_sig_untargeted_MSMICA = left_join(limma_outcome_c18neg_sig_untargeted, c18neg_MSMICA_identified_metabolites, by=c("mz"="mz", "time"="time"))

# print the results
print(limma_outcome_c18neg_sig_untargeted_MSMICA)

# save the result as a csv file
write.csv(limma_outcome_c18neg_sig_untargeted_MSMICA, "MSMICA untargeted metabolomics analysis/rcc_MSMICA_untargeted_c18neg_metabolites_limma_paired_t_test.csv", row.names = FALSE)
```
