---
title: "Case study"
author: "James Zhan"
date: "2024-07-04"
output:
  html_document: default
  pdf_document: default
---

# Outline:
- Load packages and data
- Replace the missing intensity values with half minimum imputation
- Prepare the data for MSMICA mega-scale targeted metabolomics analysis
- Prepare the data for MSMICA untargeted metabolomics analysis
- MSMICA mega-scale targeted metabolomics analysis
    - Combine the MSMICA results from HILIC positive and C18 negative renal cancer study example data (7 paired renal cancer tissue samples and normal renal tissue samples taken from the same subjects) for a limma paired t-test comparison. Half minimum imputation is used for intensites with 0 values. 
        - Reference: https://bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf
- MSMICA untargeted metabolomics analysis
    - Using HILIC positive data and C18 negative data separately, perform a limma paired t-test comparison between renal cancer tissue and normal tissue samples within the same subject, select only features with FDR-adjusted p-value <= 0.20, and use MSMICA results to annotate the selected features. Half minimum imputation is used for intensites with 0 values.
- MSMICA targeted Sparse PLS Discriminant Analysis (sPLS-DA)
    - Combine the HILIC positive and C18 negative MSMICA results for a sPLS-DA analysis to compare renal cancer tissue and normal tissue samples within the same subject, followed by the sample grouping plot (shows the relationship and similaries between sample), hierarchical clustering heatmap, and relevance network analysis using mixOmics R package. Half minimum imputation is used for intensites with 0 values.
        - Reference: http://mixomics.org/case-studies/splsda-srbct-case-study/

---

## Load packages and data
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(limma)
library(splines)
install.packages("/Users/james/Library/Mobile Documents/com~apple~CloudDocs/Desktop/Emory University - Ph.D./PhD dissertation/MSMICA/Publication/Abstract/MSMICA algorithm/MSMICA", repos = NULL, type = "source")
library(MSMICA)

setwd("/Users/james/Library/Mobile Documents/com~apple~CloudDocs/Desktop/Emory University - Ph.D./PhD dissertation/MSMICA/Publication/Abstract/MSMICA algorithm/MSMICA/case study")

# import feature table data
## hilic positive data
data(feature_table_exp_hilicpos)
print(feature_table_exp_hilicpos)
## c18 negative data
data(feature_table_exp_c18neg)
print(feature_table_exp_c18neg)

# replace all 0 values with NA after column 2
feature_table_exp_hilicpos[, -c(1:2)] = feature_table_exp_hilicpos[, -c(1:2)] %>%
    mutate_all(~na_if(., 0))

# replace all 0 values with NA after column 2
feature_table_exp_c18neg[, -c(1:2)] = feature_table_exp_c18neg[, -c(1:2)] %>%
    mutate_all(~na_if(., 0))


# median summarize the triplicates
df <- feature_table_exp_hilicpos

# Gather the triplicates into a long format
df_long <- df %>%
  pivot_longer(cols = -c(mz, time), names_to = "sample", values_to = "value")

# Extract the base sample name (without the replicate number)
df_long <- df_long %>%
  mutate(sample_base = sub("(_[0-9])$", "", sample))

# Compute the median for each set of triplicates
df_median <- df_long %>%
  group_by(mz, time, sample_base) %>%
  summarize(median_value = median(value), .groups = "drop")

# Pivot back to the wide format
feature_table_exp_hilicpos <- df_median %>%
  pivot_wider(names_from = sample_base, values_from = median_value)

# View the summarized data frame
print(feature_table_exp_hilicpos)

# median summarize the triplicates
df <- feature_table_exp_c18neg

# Gather the triplicates into a long format
df_long <- df %>%
  pivot_longer(cols = -c(mz, time), names_to = "sample", values_to = "value")

# Extract the base sample name (without the replicate number)
df_long <- df_long %>%
  mutate(sample_base = sub("(_[0-9])$", "", sample))

# Compute the median for each set of triplicates
df_median <- df_long %>%
  group_by(mz, time, sample_base) %>%
  summarize(median_value = median(value), .groups = "drop")

# Pivot back to the wide format
feature_table_exp_c18neg <- df_median %>%
  pivot_wider(names_from = sample_base, values_from = median_value)

# View the summarized data frame
print(feature_table_exp_c18neg)

# Filter out features appear less than 20% of all samples. The intensity column starts from the 3rd column.
feature_table_exp_hilicpos = QC_filter(x = feature_table_exp_hilicpos, metabolite_start_column = 3, minimum_sample_appear = 0.20)
feature_table_exp_c18neg = QC_filter(x = feature_table_exp_c18neg, metabolite_start_column = 3, minimum_sample_appear = 0.20)

# import MSMICA results
## hilic positive data
hilicpos_MSMICA_identified_metabolites = read_csv("MSMICA_test_hilicpos_MSMICA_identified_metabolites.csv")
print(hilicpos_MSMICA_identified_metabolites)
## c18 negative data
c18neg_MSMICA_identified_metabolites = read_csv("MSMICA_test_c18neg_MSMICA_identified_metabolites.csv")
print(c18neg_MSMICA_identified_metabolites)
```

## Replace the missing intensity values with half minimum imputation
```{r}

# Function to impute missing values with half of the minimum value in the row 
impute_missing_values <- function(row) {
  # Extract the values after column 2
  values <- row[-c(1, 2)]
  # Calculate half of the minimum value in the row (excluding NA)
  half_min_value <- min(values, na.rm = TRUE) / 2
  # Impute NA values with half of the minimum value
  values[is.na(values)] <- half_min_value
  # Combine the first two columns with the imputed values
  return(c(row[1:2], values))
}

# HILIC positive ############################################################################################################

# Apply the imputation function to each row
feature_table_exp_hilicpos_imputed <- t(apply(feature_table_exp_hilicpos, 1, impute_missing_values))

# Convert the result back to a data frame and ensure the column names are preserved
feature_table_exp_hilicpos_imputed <- as.data.frame(feature_table_exp_hilicpos_imputed, stringsAsFactors = FALSE)
colnames(feature_table_exp_hilicpos_imputed) <- colnames(feature_table_exp_hilicpos_imputed)
feature_table_exp_hilicpos_imputed = tibble(feature_table_exp_hilicpos_imputed)

# Round mz to 4 decimal places and time to 1 decimal place. This is because MSMICA already rounded mz to 4 decimal places and time to 1 decimal place. We need to match the feature table with MSMICA results later.
feature_table_exp_hilicpos_imputed$mz = round(feature_table_exp_hilicpos_imputed$mz, 4)
feature_table_exp_hilicpos_imputed$time = round(feature_table_exp_hilicpos_imputed$time, 1)

print(feature_table_exp_hilicpos_imputed)

# C18 negative ############################################################################################################

# Apply the imputation function to each row
feature_table_exp_c18neg_imputed <- t(apply(feature_table_exp_c18neg, 1, impute_missing_values))

# Convert the result back to a data frame and ensure the column names are preserved
feature_table_exp_c18neg_imputed <- as.data.frame(feature_table_exp_c18neg_imputed, stringsAsFactors = FALSE)
colnames(feature_table_exp_c18neg_imputed) <- colnames(feature_table_exp_c18neg_imputed)
feature_table_exp_c18neg_imputed = tibble(feature_table_exp_c18neg_imputed)

# Round mz to 4 decimal places and time to 1 decimal place. This is because MSMICA already rounded mz to 4 decimal places and time to 1 decimal place. We need to match the feature table with MSMICA results later.
feature_table_exp_c18neg_imputed$mz = round(feature_table_exp_c18neg_imputed$mz, 4)
feature_table_exp_c18neg_imputed$time = round(feature_table_exp_c18neg_imputed$time, 1)

print(feature_table_exp_c18neg_imputed)
```


## Prepare the data for MSMICA mega-scale targeted metabolomics analysis
```{r}
# Prepare the data for MSMICA mega-scale targeted metabolomics analysis ############################################################################################################
# HILIC positive ############################################################################################################
# inner join the imputed feature table and MSMICA results based on mz=mz, rt=time. This is to add intensity values to the MSMICA results.
hilicpos_MSMICA_featuretable <- inner_join(hilicpos_MSMICA_identified_metabolites, feature_table_exp_hilicpos_imputed, , by=c("mz"="mz", "time"="time"))

# remove those duplicates with different KEGGID but have the same Identified_Name
hilicpos_MSMICA_featuretable = hilicpos_MSMICA_featuretable %>%
    distinct(Identified_Name, .keep_all = TRUE)

# calculate the mean intensity for each metabolite using intensity values from all samples (starting from column 13)
hilicpos_MSMICA_featuretable$mean_intensity = rowMeans(hilicpos_MSMICA_featuretable[,13:ncol(hilicpos_MSMICA_featuretable)], na.rm = TRUE)

# move the mean_intensity column right after the identification_method column
hilicpos_MSMICA_featuretable = hilicpos_MSMICA_featuretable %>%
    relocate(mean_intensity, .after = identification_method)

# C18 negative ############################################################################################################
# inner join the imputed feature table and MSMICA results based on mz=mz, rt=time. This is to add intensity values to the MSMICA results.
c18neg_MSMICA_featuretable <- inner_join(c18neg_MSMICA_identified_metabolites, feature_table_exp_c18neg_imputed, , by=c("mz"="mz", "time"="time"))

# remove those duplicates with different KEGGID but have the same Identified_Name
c18neg_MSMICA_featuretable = c18neg_MSMICA_featuretable %>%
    distinct(Identified_Name, .keep_all = TRUE)

# calculate the mean intensity for each metabolite using intensity values from all samples (starting from column 13)
c18neg_MSMICA_featuretable$mean_intensity = rowMeans(c18neg_MSMICA_featuretable[,13:ncol(c18neg_MSMICA_featuretable)], na.rm = TRUE)

# move the mean_intensity column right after the identification_method column
c18neg_MSMICA_featuretable = c18neg_MSMICA_featuretable %>%
    relocate(mean_intensity, .after = identification_method)

# combine the HILIC positive and C18 negative MSMICA results by rows ############################################################################################################
hilicpos_c18neg_MSMICA_featuretable = rbind(hilicpos_MSMICA_featuretable, c18neg_MSMICA_featuretable)

# group by KEGGID and select the metabolites with highest mean intensity if they are the same metabolite detected in both HILIC positive and C18 negative
hilicpos_c18neg_MSMICA_featuretable_clean = hilicpos_c18neg_MSMICA_featuretable %>%
    group_by(KEGGID) %>%
    slice_max(mean_intensity) %>%
    ungroup()

# check the cleaned table
print(hilicpos_c18neg_MSMICA_featuretable_clean)

# all MSMICA identified metabolites in HILIC positive and C18 negative #############################
# transpose the table starting at column 14
hilicpos_c18neg_MSMICA_featuretable_clean_t = t(hilicpos_c18neg_MSMICA_featuretable_clean[,14:ncol(hilicpos_c18neg_MSMICA_featuretable_clean)])
hilicpos_c18neg_MSMICA_featuretable_clean_t = as.data.frame(hilicpos_c18neg_MSMICA_featuretable_clean_t)

# set the column names as the Identified_Name
colnames(hilicpos_c18neg_MSMICA_featuretable_clean_t) = hilicpos_c18neg_MSMICA_featuretable_clean$Identified_Name

# add Group_ID as a column
hilicpos_c18neg_MSMICA_featuretable_clean_t$Group_ID = rownames(hilicpos_c18neg_MSMICA_featuretable_clean_t)

# move Group_ID as the first column
hilicpos_c18neg_MSMICA_featuretable_clean_t = tibble(hilicpos_c18neg_MSMICA_featuretable_clean_t) %>%
    select(Group_ID, everything())

# Seperate the Group_ID into columns: Group, Temperature, Time, SampleID
hilicpos_c18neg_MSMICA_featuretable_clean_t = hilicpos_c18neg_MSMICA_featuretable_clean_t %>%
    separate(Group_ID, c("Group", "Temperature", "Time"), sep = "_") %>%
    # create a new column, sample_tyoe, to indicate the sample type: if Group starts with NM, then it is normal, otherwise it is cancer
    mutate(sample_type = ifelse(stringr::str_detect(Group, "^NM"), "normal", "cancer")) %>%
    # move sample type to the first column
    select(sample_type, everything())
    

# separate the Group column into Group and ParticipantID: the first two characters are Group, the rest are ParticipantID
hilicpos_c18neg_MSMICA_featuretable_clean_t = hilicpos_c18neg_MSMICA_featuretable_clean_t %>%
    separate(Group, c("Group", "ParticipantID"), sep = 2)

# Convert all intensity values to log2
hilicpos_c18neg_MSMICA_featuretable_clean_t[, -c(1:5)] = hilicpos_c18neg_MSMICA_featuretable_clean_t[, -c(1:5)] %>%
    mutate_all(~log2(.))

# check the data ready for MSMICA mega-scale targeted metabolomics analysis
print(hilicpos_c18neg_MSMICA_featuretable_clean_t)
```


## Prepare the data for MSMICA untargeted metabolomics analysis
```{r}
# Prepare the data for MSMICA untargeted metabolomics analysis ############################################################################################################
# HILIC positive ############################################################################################################
# all MSMICA identified metabolites in HILIC positive and C18 negative #############################
# transpose the table starting at column 13
feature_table_exp_hilicpos_imputed_t = t(feature_table_exp_hilicpos_imputed[,3:ncol(feature_table_exp_hilicpos_imputed)])
feature_table_exp_hilicpos_imputed_t = as.data.frame(feature_table_exp_hilicpos_imputed_t)

# set the column names as the combination of mz and time
colnames(feature_table_exp_hilicpos_imputed_t) = paste0(feature_table_exp_hilicpos_imputed$mz, "_", feature_table_exp_hilicpos_imputed$time)

# add Group_ID as a column
feature_table_exp_hilicpos_imputed_t$Group_ID = rownames(feature_table_exp_hilicpos_imputed_t)

# move Group_ID as the first column
feature_table_exp_hilicpos_imputed_t = tibble(feature_table_exp_hilicpos_imputed_t) %>%
    select(Group_ID, everything())

# Seperate the Group_ID into columns: Group, Temperature, Time, SampleID
feature_table_exp_hilicpos_imputed_t = feature_table_exp_hilicpos_imputed_t %>%
    separate(Group_ID, c("Group", "Temperature", "Time"), sep = "_") %>%
    # create a new column, sample_tyoe, to indicate the sample type: if Group starts with NM, then it is normal, otherwise it is cancer
    mutate(sample_type = ifelse(stringr::str_detect(Group, "^NM"), "normal", "cancer")) %>%
    # move sample type to the first column
    select(sample_type, everything())
    
# separate the Group column into Group and ParticipantID: the first two characters are Group, the rest are ParticipantID
feature_table_exp_hilicpos_imputed_t = feature_table_exp_hilicpos_imputed_t %>%
    separate(Group, c("Group", "ParticipantID"), sep = 2)

# Convert all intensity values to log2
feature_table_exp_hilicpos_imputed_t[, -c(1:5)] = feature_table_exp_hilicpos_imputed_t[, -c(1:5)] %>%
    mutate_all(~log2(.))

# check the data ready for MSMICA untargeted metabolomics analysis
print(feature_table_exp_hilicpos_imputed_t)

# C18 negative ############################################################################################################
# all MSMICA identified metabolites #############################
feature_table_exp_c18neg_imputed_t = t(feature_table_exp_c18neg_imputed[,3:ncol(feature_table_exp_c18neg_imputed)])
feature_table_exp_c18neg_imputed_t = as.data.frame(feature_table_exp_c18neg_imputed_t)

# set the column names as the combination of mz and time
colnames(feature_table_exp_c18neg_imputed_t) = paste0(feature_table_exp_c18neg_imputed$mz, "_", feature_table_exp_c18neg_imputed$time)

# add Group_ID as a column
feature_table_exp_c18neg_imputed_t$Group_ID = rownames(feature_table_exp_c18neg_imputed_t)

# move Group_ID as the first column
feature_table_exp_c18neg_imputed_t = tibble(feature_table_exp_c18neg_imputed_t) %>%
    select(Group_ID, everything())

# Seperate the Group_ID into columns: Group, Temperature, Time, SampleID
feature_table_exp_c18neg_imputed_t = feature_table_exp_c18neg_imputed_t %>%
    separate(Group_ID, c("Group", "Temperature", "Time"), sep = "_") %>%
    # create a new column, sample_tyoe, to indicate the sample type: if Group starts with NM, then it is normal, otherwise it is cancer
    mutate(sample_type = ifelse(stringr::str_detect(Group, "^NM"), "normal", "cancer")) %>%
    # move sample type to the first column
    select(sample_type, everything())

# separate the Group column into Group and ParticipantID: the first two characters are Group, the rest are ParticipantID
feature_table_exp_c18neg_imputed_t = feature_table_exp_c18neg_imputed_t %>%
    separate(Group, c("Group", "ParticipantID"), sep = 2)

# Convert all intensity values to log2
feature_table_exp_c18neg_imputed_t[, -c(1:5)] = feature_table_exp_c18neg_imputed_t[, -c(1:5)] %>%
    mutate_all(~log2(.))

# check the data ready for MSMICA untargeted metabolomics analysis
print(feature_table_exp_c18neg_imputed_t)
```

## MSMICA mega-scale targeted metabolomics analysis
```{r}
############################# limma paired t-test #############################
# Factorize the samples into normal and cancer
Group = factor(hilicpos_c18neg_MSMICA_featuretable_clean_t$sample_type, levels = c("normal", "cancer"))

# Factorize the ParticipantID
ParticipantID = factor(hilicpos_c18neg_MSMICA_featuretable_clean_t$ParticipantID)

# Create the design matrix
design_T <- model.matrix(~ ParticipantID+Group)

# Extract the intensity values for the limma analysis
hilicpos_c18neg_intensity = hilicpos_c18neg_MSMICA_featuretable_clean_t[, 6:ncol(hilicpos_c18neg_MSMICA_featuretable_clean_t)]
hilicpos_c18neg_intensity = as.data.frame(hilicpos_c18neg_intensity)

# transpose the metabolite intensity data
y_T = t(hilicpos_c18neg_intensity)

# Fit the linear model
fit = lmFit(y_T, design_T)
fit2 <- eBayes(fit)

# chcek the top table in the limma paited t-test output
topTable(fit2, coef="Groupcancer", adjust="BH")

# Extract all relevant values from the limma output
limma_outcome_1 = data.frame(
    Cancer_log2FC = c(fit2$coefficients[, "Groupcancer"]),
    t_statistic = c(fit2$t[, "Groupcancer"]), 
    p = c(fit2$p.value[, "Groupcancer"]))

limma_outcome_1$p.fdr = p.adjust(limma_outcome_1$p, method="BH")
limma_outcome_1$metabolite = colnames(hilicpos_c18neg_MSMICA_featuretable_clean_t[, -c(1:5)])

# add KEGGID, Exact_mass, Adduct, mz, time from hilicpos_c18neg_MSMICA_featuretable_clean to limma_outcome_1
limma_outcome_1$KEGGID = hilicpos_c18neg_MSMICA_featuretable_clean$KEGGID
limma_outcome_1$Exact_mass = hilicpos_c18neg_MSMICA_featuretable_clean$Exact_mass
limma_outcome_1$Adduct = hilicpos_c18neg_MSMICA_featuretable_clean$Adduct
limma_outcome_1$mz = hilicpos_c18neg_MSMICA_featuretable_clean$mz
limma_outcome_1$time = hilicpos_c18neg_MSMICA_featuretable_clean$time

# Organize the column names: KEGGID, metabolite, Exact_mass, Adduct, mz, time, t_statistics, p, p.fdr, Cancer_log2FC
limma_outcome_1 = limma_outcome_1 %>%
    select(KEGGID, metabolite, Exact_mass, Adduct, mz, time, t_statistic, p, p.fdr, Cancer_log2FC) %>%
    arrange(p) %>%
    tibble()

# save the significant metabolites
limma_outcome_1_sig = limma_outcome_1 %>%
    filter(p < 0.05)

# print the results
print(limma_outcome_1_sig, width=Inf)

# create a new folder called: MSMICA mega-scale targeted metabolomics analysis
dir.create("MSMICA mega-scale targeted metabolomics analysis")

# save the result as a csv file
write.csv(limma_outcome_1, "MSMICA mega-scale targeted metabolomics analysis/rcc_targeted_MSMICA_HILICposC18neg_metabolites_limma_paired_t_test.csv", row.names = FALSE)
```


## MSMICA untargeted metabolomics analysis
```{r}
############################# limma paired t-test #############################
# HILIC positive ############################################################################################################
# Factorize the samples into normal and cancer
Group_hilicpos = factor(feature_table_exp_hilicpos_imputed_t$sample_type, levels = c("normal", "cancer"))

# Factorize the ParticipantID
ParticipantID_hilicpos = factor(feature_table_exp_hilicpos_imputed_t$ParticipantID)

# Create the design matrix
design_T_hilicpos <- model.matrix(~ ParticipantID_hilicpos+Group_hilicpos)

# Extract the intensity values for the limma analysis
hilicpos_intensity = feature_table_exp_hilicpos_imputed_t[, 6:ncol(feature_table_exp_hilicpos_imputed_t)]
hilicpos_intensity = as.data.frame(hilicpos_intensity)

# transpose the metabolite intensity data
y_T_hilicpos = t(hilicpos_intensity)

# Fit the linear model
fit_hilicpos = lmFit(y_T_hilicpos, design_T_hilicpos)
fit2_hilicpos <- eBayes(fit_hilicpos)

# chcek the top table in the limma paited t-test output
topTable(fit2_hilicpos, coef="Group_hilicposcancer", adjust="BH")

# Extract all relevant values from the limma output
limma_outcome_hilicpos = data.frame(
    Cancer_log2FC = c(fit2_hilicpos$coefficients[, "Group_hilicposcancer"]),
    t_statistic = c(fit2_hilicpos$t[, "Group_hilicposcancer"]), 
    p = c(fit2_hilicpos$p.value[, "Group_hilicposcancer"]))

limma_outcome_hilicpos$p.fdr = p.adjust(limma_outcome_hilicpos$p, method="BH")
limma_outcome_hilicpos$mz_time = colnames(feature_table_exp_hilicpos_imputed_t[, -c(1:5)])
limma_outcome_hilicpos = tibble(limma_outcome_hilicpos)

# mz, time from feature_table_exp_hilicpos_imputed to limma_outcome_hilicpos
limma_outcome_hilicpos$mz = feature_table_exp_hilicpos_imputed$mz
limma_outcome_hilicpos$time = feature_table_exp_hilicpos_imputed$time

# Organize the column names: mz, time, t_statistics, p, p.fdr, Cancer_log2FC
limma_outcome_hilicpos = limma_outcome_hilicpos %>%
    select(mz, time, t_statistic, p, p.fdr, Cancer_log2FC) %>%
    arrange(p) %>%
    tibble()

# select only the significant metabolites
limma_outcome_hilicpos_sig = limma_outcome_hilicpos %>%
    filter(p < 0.05)

# left join the MSMICA results to the limma_outcome_hilicpos_sig based on mz and time
limma_outcome_hilicpos_sig_MSMICA = left_join(limma_outcome_hilicpos_sig, hilicpos_MSMICA_identified_metabolites, by=c("mz"="mz", "time"="time"))

# print the results
print(limma_outcome_hilicpos_sig_MSMICA, n=20, width=Inf)

# create a new folder called: MSMICA mega-scale targeted metabolomics analysis
dir.create("MSMICA untargeted metabolomics analysis")

# save the result as a csv file
write.csv(limma_outcome_hilicpos_sig_MSMICA, "MSMICA untargeted metabolomics analysis/rcc_MSMICA_untargeted_hilicpos_metabolites_limma_paired_t_test.csv", row.names = FALSE)

# C18 negative ############################################################################################################
# Factorize the samples into normal and cancer
Group_c18neg = factor(feature_table_exp_c18neg_imputed_t$sample_type, levels = c("normal", "cancer"))

# Factorize the ParticipantID
ParticipantID_c18neg = factor(feature_table_exp_c18neg_imputed_t$ParticipantID)

# Create the design matrix
design_T_c18neg <- model.matrix(~ ParticipantID_c18neg+Group_c18neg)

# Extract the intensity values for the limma analysis
c18neg_intensity = feature_table_exp_c18neg_imputed_t[, 6:ncol(feature_table_exp_c18neg_imputed_t)]

# transpose the metabolite intensity data
y_T_c18neg = t(c18neg_intensity)

# Fit the linear model
fit_c18neg = lmFit(y_T_c18neg, design_T_c18neg)
fit2_c18neg <- eBayes(fit_c18neg)

# chcek the top table in the limma paited t-test output
topTable(fit2_c18neg, coef="Group_c18negcancer", adjust="BH")

# Extract all relevant values from the limma output
limma_outcome_c18neg = data.frame(
    Cancer_log2FC = c(fit2_c18neg$coefficients[, "Group_c18negcancer"]),
    t_statistic = c(fit2_c18neg$t[, "Group_c18negcancer"]), 
    p = c(fit2_c18neg$p.value[, "Group_c18negcancer"]))

limma_outcome_c18neg$p.fdr = p.adjust(limma_outcome_c18neg$p, method="BH")
limma_outcome_c18neg$mz_time = colnames(feature_table_exp_c18neg_imputed_t[, -c(1:5)])
limma_outcome_c18neg = tibble(limma_outcome_c18neg)

# mz, time from feature_table_exp_c18neg_imputed to limma_outcome_c18neg
limma_outcome_c18neg$mz = feature_table_exp_c18neg_imputed$mz
limma_outcome_c18neg$time = feature_table_exp_c18neg_imputed$time

# Organize the column names: mz, time, t_statistics, p, p.fdr, Cancer_log2FC
limma_outcome_c18neg = limma_outcome_c18neg %>%
    select(mz, time, t_statistic, p, p.fdr, Cancer_log2FC) %>%
    arrange(p) %>%
    tibble()

# select only the significant metabolites
limma_outcome_c18neg_sig = limma_outcome_c18neg %>%
    filter(p < 0.05)

# left join the MSMICA results to the limma_outcome_c18neg_sig based on mz and time
limma_outcome_c18neg_sig_MSMICA = left_join(limma_outcome_c18neg_sig, c18neg_MSMICA_identified_metabolites, by=c("mz"="mz", "time"="time"))

# print the results
print(limma_outcome_c18neg_sig_MSMICA, n=20, width=Inf)

# save the result as a csv file
write.csv(limma_outcome_c18neg_sig_MSMICA, "MSMICA untargeted metabolomics analysis/rcc_MSMICA_untargeted_c18neg_metabolites_limma_paired_t_test.csv", row.names = FALSE)
```

## MSMICA targeted Sparse PLS Discriminant Analysis (sPLS-DA)
```{r, fig.width=10, fig.height=10, fig.fullwidth=TRUE}
library(readr)
library(dplyr)
library(tidyr)
library(mixOmics)

set.seed(2024) # for reproducibility, remove for normal use

# This is the intensity data for the MSMICA identified metabolites in HILIC positive and C18 negative
X = as.matrix(hilicpos_c18neg_intensity)

# This is the sample type for each sample: normal or cancer
Y = factor(hilicpos_c18neg_MSMICA_featuretable_clean_t$sample_type, levels = c("normal", "cancer"))

dim(X) # check the dimensions of the X dataframe
length(Y) # check the length of the Y dataframe

# Preliminary Analysis with PCA
# run pca method on data (our intensity data was already log2 transformed, so we don't need to scale it here)
pca.RCC = pca(X, ncomp = 10, center = TRUE, scale = FALSE) 
plot(pca.RCC)  # barplot of the eigenvalues (explained variance per component)

# Two components would be sufficient to explain a moderate proportion of the data's variance
plotIndiv(pca.RCC, group = Y, ind.names = TRUE, # plot the samples projected
          legend = TRUE, 
          col.per.group = c("#F68B33", "#388ECC"), # set of colours
          title = 'PCA on the RCC study, comp 1 - 2') # onto the PCA

# It seems like only the normal tissue samples are separated from the cancer tissue samples. Cancer tissue samples are heterogenous. This is expected as the normal tissue samples are expected to have different metabolite profiles compared to the cancer tissue samples.

# Sparse PLS-DA - Note, sPLS-DA is a supervised method, so we need to provide the class labels (Y) to the function. This is designed to find the metabolites that best discriminate between the two classes. Thus, it is not surprising that the sPLS-DA model will perform better than the PCA model for class discrimination.
RCC.splsda <- splsda(X, Y, ncomp = 10)  # set ncomp to 10 for performance assessment later

# Tuning sPLS-DA: Selecting the number of components
# Our sample size is too low that even two fold cross-validation had a error. Thus, we just select the first two components for the sPLS-DA model arbitrarily.

# Tuning sPLS-DA: Selecting the number of variables
# grid of possible keepX values that will be tested for each component
list.keepX <- c(5:10,  seq(10, 200, 10))

# undergo the tuning process to determine the optimal number of variables
tune.splsda.RCC <- tune.splsda(X, Y, ncomp = 2, # calculate for first 2 components
                                 validation = 'Mfold',
                                 folds = 5, nrepeat = 10, # use repeated cross-validation
                                 dist = 'max.dist', # use max.dist measure
                                 measure = "BER", # use balanced error rate of dist measure
                                 test.keepX = list.keepX,
                                 cpus = 1) # allow for paralleliation to decrease runtime

# Since at least one component is needed, we manually set the optimal number of components to 2 for the sPLS-DA model, though the optimal number of components is 1 according to the tuning process.
optimal.ncomp <- 2 # what is the optimal value of components according to tune.splsda()
optimal.keepX <- tune.splsda.RCC$choice.keepX # what are the optimal values of variables according to tune.splsda()

print(optimal.ncomp) # print the optimal number of components
print(optimal.keepX) # print the optimal number of variables

# Final Model for sPLS-DA
# fit the final sPLS-DA model with the optimal number of components and variables
final.splsda <- splsda(X, Y, 
                       ncomp = optimal.ncomp, 
                       keepX = optimal.keepX)

# plot the samples projected onto the first two components of the sPLS-DA subspace
plotIndiv(final.splsda, comp = c(1,2), # plot samples from final model
          group = Y, ind.names = TRUE, # colour by class label
          ellipse = TRUE, legend = TRUE, # include 95% confidence ellipse,
          col.per.group = c("#F68B33", "#388ECC"), # set of colours
          title = 'Final sPLS-DA on SRBCT, comp 1 & 2')

# hierarchical clustering heatmap
# set the styling of the legend to be homogeneous with previous plots
legend=list(legend = levels(Y), # set of classes
            col = unique(color.mixo(Y)), # set of colours
            title = "Sample Type", # legend title
            cex = 0.7) # legend size

# generate the CIM, using the legend and colouring rows by each sample's class
cim(
    final.splsda, 
    ## add the participant clustering color
    row.sideColors = color.mixo(Y), 
    ## remove the participant values
    row.names = FALSE,
    ## increase the margins
    margin = c(5, 25),
    ## make the text bigger
    col.cex = 1,
    legend = NULL, 
    transpose = TRUE)

# Add the legend manually
legend("topright", legend = legend$legend, fill = legend$col, title = legend$title)

# network analysis:
network(final.splsda, color.node = c("orange","lightblue"), cutoff = 0.63, size.node = 0.5, cex.node.name = 0.7, cex.edge.label = 0.7, symkey = TRUE) 
```


